---
title: "R crash course / 02"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    widescreen: true
    smaller: true
    css: files/style.css
    logo: files/logo_bihealth_en.png
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

## Aims for today

* Complex data structures (matrices, lists and data frames)
* Functions in R
* Reading data from files

## Repetitorium of Day 1

Questions, anybody?

## Functions in R: a brief introduction

```{r eval=FALSE,echo=TRUE}
samples   <- c(1, 10, 23, 42, 13)
samples_n <- length(samples)
```

 * Both `c` and `length` are functions. They take some arguments (often many
of them) and return a single object: a vector, a matrix or something else.

 * You can always assign the result of the function to a variable.

 * Sometimes functions return `NULL`, which is R for "nothing", but which is
still something you can use or assign.


## Exercise 2.1 More on vectors (and functions)

There is a lake in a garden. Every day, the water lilies cover twice as
much area as the previous day. On the first day, the water lilies cover
1/100th of the area of the lake.

 * What is the formula to calculate the area covered by the water lilies on
   day n? (pen and paper, not in Rstudio!)
 * assign days 1 ... 10 to variable x (use `c()` or `:`). Now assign the fraction of the area
   covered by water lilies on day n to variable y
 * plot x vs y (use the simple `plot()` function)
 * When will the half of the area be covered by water lilies? Use `abline(h=.5)`
   (what does it do?) to show a graphical solution
 * What is the fraction on day 3? Hint: what does `y[3]` do?
 * Speaking of which, what does `y[4:5]` do? (just try it!)

## Exercise 2.2: More on plotting

 * Make the plot yourself.
 * use 'col' parameter of `plot` and a color name (e.g. "red") to change
   the color of the line (`plot(..., col="red")`)
 * what if each day the lilies cover a fraction of area that is 1.5 the
   fraction of area on the previous day?
 * Use `lines(x, y)` function to put a second line on the plot
 * For more experienced users: Create a Rmarkdown.

## Logical vectors

In R, there are two special values: `TRUE` and `FALSE`. They can be used to
create logical vectors.

```{r eval=FALSE, echo=TRUE}
sel     <- c(TRUE, TRUE, TRUE, TRUE, FALSE)
sel
!sel
```

Comparison operators (`>`, `<`, `<=`, `>=`, `==`, `!=`) produce logical
vectors:

```{r eval=FALSE, echo=TRUE}
samples <- c(1, 1, 2, 5, 7)
samples > 2
which(samples == 7)
which(samples != 1)
```


## Logical vectors

Logical vectors can be used to access elements:

```{r eval=FALSE,echo=TRUE}
persons <- c("Aphrodite", "Bacchus", "Circe", "Demeter", "Eurypides")
sel     <- c(TRUE, TRUE, TRUE, TRUE, FALSE)
persons[sel]

# we can abbreviate the TRUE and FALSE to T and F (avoid)
greek <- persons[ c(T, F, T, T, T) ]
```



## Exercise 2.3: more on vectors and functions, introducing NA

Create a vector as follows:


```{r echo=TRUE,eval=FALSE}
samples <- c(1, 10, NA, 15)
```

`NA` stands for `not available` (e.g., missing data)

 * try adding a number to that vector
 * what does `length(samples)` return?
 * what does `mean(samples)` return? Why is that?
 * use the **named parameter** `na.rm=TRUE` for the `mean()` function. Look
   up help (`?mean`) to see how it can be used. What happens now?
 * what does the `is.na()` function return when applied to `samples`?
 * how do you find `NA` values? Use `is.na` and `which`
 * How do you select only these samples which are not `NA`?



## Extra (if you are bored): Creating a function

The reason we are showing how to create a function is to show you that it
is simple, and also because it will help you understand what functions are.

```{r eval=FALSE,echo=TRUE}
#' Function name
#' Function description
some_name <- function(param1, param2=2) {

## code comment
# <your code goes in here>

}
```

 * If you have to write something more than twice, you might want to turn it into
   a function!
 * R is a *functional* language:
    * almost everything is a function
    * functions usually take something and return a value; they are **not**
      supposed to use or modify anything that they have not been passed to
      explicitely (sometimes hard to avoid)


# Complex data structures

## Main data representations in R

 * vectors (string, number, integer, logic, factor)
 * matrices and arrays
 * lists
 * data frames

## Data matrices

Much like vectors, matrices can only hold one data type (e.g. only numeric
or only character or only logical etc.).

```{r}
m <- matrix(1:18, ncol=3, nrow=6)
# compare with
m <- matrix(1:18, ncol=3, nrow=6, byrow=TRUE)
dim(m)
ncol(m)
nrow(m)
```

## Accessing columns, rows and element

`matrix[row, column]`

So, for example:

```{r}
m[1, ] # vector which is the first row
m[, 2] # vector which is the first column
m[3, 1] # first element of the third row
```

## Row and column names

We can name rows and columns of a matrix and use the names to access the
rows and columns:

```{r}
colnames(m) <- letters[1:ncol(m)]
rownames(m) <- LETTERS[1:nrow(m)]

m["A", "b"]
m["B", ] # etc.
```

## Remember

 * If you select a single column or a single row, you will get a vector
 * If you select more than one row or column, you will get a (smaller) matrix
 * If you select more rows or columns than are present, you will get a
   "subscript out of bonds" error

## Demonstration


## Lists

 * Lists hold all types of information
 * Lists are really, really cool
 * Lists have elements. An element of a list can have *any* type, including
   another list.

## Lists

 Caveats:

 * You access elements of a list using `[[`, not `[`
 * Lists may have names (set with `names()`), but don't have to

## Data frames

Data frames are a bit like matrices, but every column can store different
type of data. In this, they are more like lists (which they in fact are).

```{r}
names <- c("January", "Manuela", "Bill")
lastn <- c("Weiner", "Benary", "Gates")
age   <- c(1001, NA, 65)

d <- data.frame(names=names, last_names=lastn, age=age)
class(d)
class(d[,1])
class(d[,3])
```

## Accessing elements in data frames

You can access the data frame elements much like the elements of a matrix.

However, since data frames are lists, the list operator (`$`) also works:

```{r}
d$names # same as d[,1] or d[, "names"]
d$lastn
d$lastn[1]
```

However, note that when you select a row, you will get a data frame, not a
vector. This is because each of the column can be of different type, and
vectors can hold only one type of data.

## Matrices vs data frames

 Caveats:

  * data frames sometimes turn strings to factors (more on that later), which may have really
    disastrous consequences, use `stringsAsFactors=FALSE`
  * small typos can turn a numeric vector into a string
  * factors are dangerous to work with, use them cautiously

  Gory details: matrices are a basic data type. Data frames are a list.

## Data frames vs tibble

 Caveats:

 * tibbles are the data frames from tidyverse
 * Whatever you can do to a data frame, you can do to a tibble as well
 * `read_*` functions return a tibble
 * tibble do not have row names
 * If you select a single row in a data frame, you get a smaller data
   frame. If you select a single column, you get a vector.
 * In tibble, you always get a smaller tibble.


## Exercise 2.4

 * Create a 5x3 matrix with random numbers. Use `matrix` and `rnorm`.
 * Turn the matrix into a data frame. Use `as.data.frame` for that.
 * Add column and row names.
 * Add a column. Each value in the column should be "A" (a string). Use the
   `rep` function for that.
 * Add a column with five numbers from 0 to 1. Use the `seq` function for that.

## Using rep

`rep()` is used to replicate vectors.

```{r}
rep(c("A", "B"), 5) 
# result:
#  [1] "A" "B" "A" "B" "A" "B" "A" "B" "A" "B"

rep(c("A", "B"), each=5)
# result
# [1] "A" "A" "A" "A" "A" "B" "B" "B" "B" "B"
```

# Reading and writing data

## Reading data

Main data types you will encounter:

---------------------------- ------------------------------- ------------------------------- --------------------------
Data type                    Function                        Package                         Notes
---------------------------- ------------------------------- ------------------------------- --------------------------
Columns separated by spaces  `read_table()`                  `readr`                         one or more
                                                                                             spaces separate 
                                                                                             each column
                                                             
TSV / TAB separated values   `read_tsv()`                    `readr`                         Delimiter is tab (`\t`).
                                                             
CSV / comma separated        `read_csv()`                    `readr`                         Comma separated values
                                                                                             
Any delimiter                `read_delim()`                  `readr`                         Customizable
                                                             
XLS (old Excel)              `read_xls()`                    `readxl`                        Just don't use it.
                             `read_excel()`                                                  From the
                                                                                             `readxl`
                                                                                             package.
                                                             
XLSX (new Excel)             `read_xlsx()`                   `readxl`                        From the `readxl` package.
                             `read_excel()`                                                  You need to
                                                                                             provide the
                                                                                             sheet number
                                                                                             you wish to
                                                                                             read. 
                                                                                             Note: returns a tibble,
                                                                                             not a data frame! 
---------------------------- ------------------------------- ------------------------------- --------------------------

Note: there are also "base R" functions `read.table`, `read.csv`,
`read.tsv` (there is no function for reading XLS[X] files in base R). The
tidyverse functions above are preferable.


## Excercise 2.5

Read, inspect the following files:

 * `TB_ORD_Gambia_Sutherland_biochemicals.csv`
 * `iris.csv`
 * `meta_data_botched.xlsx` 

 1. Which functions would you use? 
 1. What kind of issues can you detect?
 1. Suggestions of solving the issues?
 
The function `readxl_example("deaths.xls")` returns a file name. Read this file. How can you omit the lines at the top and at the bottom of the file?
(hint: `?read_excel`). How can you force the date columns to be interpreted
as dates and not numbers?


## Tibbles / readxl

tibbles belong to the tidyverse. They are nice to work with and very
useful, but we can stick to data frames for now. Therefore, do

```{r}
mydataframe <- as.data.frame(read_xlsx("file.xlsx"))
```

One crucial difference between tibble and data frame is that `tibble[ , 1 ]` 
returns a tibble, while `dataframe[ , 1]` returns a vector. The second
crucial difference is that it does not support row names (on purpose!).



