---
title: "R crash course / 03"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    widescreen: true
    smaller: true
    css: files/style.css
    logo: files/logo_bihealth_en.png
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

## Aims for today

* Cleaning data
* Regular expressions
* Searching, sorting and selecting

## Repetitorium of Day 2

# Cleaning data
## A short note on organizing your data

 * If possible, use a specialized tool and not Excel (a database,
   specialized clinical data management system, REDCap etc.)

## And if you use Excel...

 If you use Excel, follow these rules:

   * NEVER modify original files
   * Versionize your files
   * Check data for consistency (e.g. with R!)
   * Use preselected values 
   * Add a meta-data sheet describing the fields
   * do not color-code or format-code your information (e.g. denoting treatment in bold)
   * Add comments in separate fields
   * NEVER throw away or replace any identifiers, always keep the old identifiers
   * understand what you describe (patients? samples? assays?)
   * understand the difference between wide and long format

## And if you use Excel and don't follow these rules

  * brace yourself for bad times
  * get used to nasty remarks

## Diagnosing problems

Potential problems:

 * incorrect import (e.g. numeric values as strings)
 * typos (e.g. "x15" instead of "15")
 * incongruent labels (e.g. mixing "M", "male" and "männlich")
 * mixing upper and lower case
 * mixing different types of decimal separator
 * special numeric values (e.g. "> 10", "< 5")
 * comments in the same column ("10 (unreliable)")
 * dates (also, Excel converts certain strings to dates)

## Diagnosing problems

 * tidyverse reading functions provide a summary on the reading process,
   e.g.:


```{r eval=TRUE,results="markdown"}
library(tidyverse)
myiris <- read_csv("../Datasets/iris.csv")
```

 * Use `summary`, `class` and `typeof` functions to find potential
   problems. This is what we expect:


```{r eval=TRUE,results="markdown"}
typeof(myiris$Petal.Width)
class(myiris$Petal.Width)
summary(myiris$Petal.Width)
```

Longer explanation: `typeof` returns the so called "storage type", that is
the very basic type of a value (string, logical, integer, list etc.).
`class` can be different (more specific). For example, if `mtx` is a matrix
of numbers, `typeof(mtx)` is `double`, and `class(mtx)` is `matrix`.

## Diagnosing problems

 * Use `summary`, `class` and `typeof` functions to find potential
   problems. This is **not** what we expect:

```{r eval=TRUE,results="markdown"}
 typeof(myiris$`Sepal Length`)
  class(myiris$`Sepal Length`)
summary(myiris$`Sepal Length`)
```

(we use the back ticks because the column name contains a space)




## Diagnosing problems

 * The colorDF package provides a function called `summary_colorDF` which
   can be used to diagnose problems with different flavors of data frames:





```{r eval=TRUE,results="markdown",R.options=list(width=100)}
library(colorDF)
summary_colorDF(myiris)
```

## Mending problems

 * find the incorrect values
 * replace incorrect values (by search and replace)
 * enforcing a particular data type (e.g., converting strings to numbers)


## Substitutions (search & replace)

 * `gsub(string1, string2, text)` substitute all occurences of `string1` in
    by `string2` in `text`
 * `sub(...)` same, but only the first occurence

```{r}
## Clearly broken!
group <- c("ontrol", "Montrol", "Kontrol", "Kkkkkontrl", "hello Connnnnnntrol", "oh what a nice day it is, controooool", "control ", "Control, incoming")

## replace upper-case "Control"
group2 <- gsub("[ckCK]+on*tro*l$", "control", group)

grepl("[ckCK]+on*tro*l$", group)

group != group2

## replace all spaces
group <- gsub(" ", "", group)
```

Notes:

The above operations are so common that there are "shortcut" functions
defined in R:

```{r}
## get rid of the spaces
trimws(group)

## make everything lower case
tolower(group)

## make everything upper case
toupper(group)
```


## Regular expressions

Is there not a way to simplify things? 

```{r}
group <- c("Control", " control", "control ", "Control   ")

# one method
gsub(" ", "", group) # but what happens if the groups are "group 1", "group 2" etc?

# substitute at the end of the string
gsub(" $", "", group)

# at the beginning
gsub("^ ", "", group)

# any number of occurences
gsub(" *", "", group)
```

```{r}
group <- c("Control", " control", "control ", "Control   ")
group1 <- trimws(group)
group2 <- tolower(group1)

## sin(log(tan(pi)))
group <- tolower(trimws(group))

```



## Regular expression cont.

In regular expressions, some characters or strings have special meanings

 * Some mean what is supposed to be replaced (e.g. `.`, dot,
   means "any character")
 * Some mean how many times something gets matched; e.g. `*` means "any
   number of matches of the previous character, so `.*` means any character
   any number of times (like, everything)
 * Some characters at certain position may also have a special meaning
   (like `^` means "match the beginning of a string" and `$` at the end of
   a regular expression means "match at the end of a string").
 * A string like `[abc]` will match only a, b and c, but not d. You can use
   ranges, e.g. `[0-9]` or `[a-z]`

## Regular expression cont.

```{r}
foo <- c("a1_123", "a1_231", "a2_17", "B2_23", "c2_889")

# substitute digits only
gsub("[0-9]*", "", foo)

# substitute letters only
gsub("[a-z]*", "", foo)

# same, but case insensitive
gsub("[a-z]*", "", foo, ignore.case=TRUE)

# substitute everything after "_"
# dot means "any character"
gsub("_.*", "", foo)
```

## Regular expression cont.

Note: the function `gsub` takes three arguments:

 * a regular expression
 * a replacement
 * a character vector

It then finds all occurences of the regular expression in each element of
the character vector and replaces them by the replacement.

## Typical example

Often, due to typos, when you read a data file, a numeric column becomes a
character string, because for example someone entered "x15" instead of "15"
or mixed German and English decimal separators (e.g. "1,5" and "1.5" in the
same column).

We can use the `as.numeric` function to convert the strings to numeric
values, but this does not work as intended:


```{r eval=TRUE, results="markdown"}
foo <- c("xxxxx15", "15.5", "15,5")
as.numeric(foo)
```

We need to substitute the strings first:

```{r eval=TRUE, results="markdown"}
foo <- gsub("[a-zA-Z]*", "", foo) # remove letters
foo <- gsub(",", ".", foo)        # convert decimal separator
as.numeric(foo)
```



## Regexps cont.

You can wrap gsub around another gsub

```{r}
vec <- c("mouse", " Mouse", "mus domesticus", "chicken", "Schicken", "Duck")
vec <- gsub("S*chicken", "Chicken", gsub("^m.*", "Mouse", gsub("^ *", "", vec)))

## same as

vec <- gsub("^ *", "", vec)
vec <- gsub("^m.*", "Mouse", vec)
vec <- gsub("S*chicken", "Chicken", vec)
```

## Exercise 3.1


 * What regular expression matches all of the ankyrin repeat genes (but not other genes) in the following vector: `c("ANKRD22", "ANKEN", "ank.rep.domain 12", "ifng-1", "ANKRD-33", "  ankrd23", "MAPK)`
 * Use gsub in the above vector to replace all occurences of "ANKRD"   
 * Use gsub to make the following uniform: `c("male", "Male ", "M", "F", "female", " Female")`
 * Using `gsub` and the `toupper` function, clean up the gene names such
   that they conform to the HGNC (all capital
   letters, no spaces, no dashes): `c("ankrd22", "ifng", "Nf-kb", " Cxcl 5", "CCL 6.", "ANK.r.d. 12")`


## Using regular expressions to clean tables 


## Exercise 3.2

 * Read the data file `iris.csv` using the `read_csv` function. Spot and correct errors.
 * Read the data file `meta_data_botched.xlsx`. Spot and correct errors.

## Writing data

Main data types you will encounter:

---------------------------- ------------------------------- ------------------------------- --------------------------
Data type                    Function                        Package                         Notes
---------------------------- ------------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `write.table()`                 `base R`                        Tab is `\t`.
                                                                                             A very general and 
                                                                                             customizable function
                                                                                             Warning:
                                                                                             column names 

                             `write_tsv()`                   `readr`                         No rownames!

                                                                                             
CSV / comma separated        `write.csv()`                   `base R`                        a wrapper around
                                                                                             `write.table()`
                                                                                             
                             `write_csv()`                   `readr`                         No rownames!

XLS (old Excel)                                                                              Just don't use it. No, 
                                                                                             seriously, don't.
                                                                                             
XLSX (new Excel)             `write_xlsx()`                  `writexl`                       Warning:
                                                                                             row names
                                                                                             omitted!
---------------------------- ------------------------------- ------------------------------- --------------------------


## Row names – gory details.

The default base R functions `read.*` assign (unless otherwise specified)
row names to a data frame.

The tidyverse `read_*` family of functions return not a data frame, but its
variant, a *tibble*. These tibbles behave very much like data frame;
however, **tidyverse discourages the use of row names** and sometimes even
removes existing row names on data frames.

## Wide and Long format (demonstration)


## Wide and Long format

 Long advantages:

  * easier to filter, process, visualize, do statistics with
  * focused on measurement ("patient ID" or equivalent is a covariate, and so is measurement type)
 
 Wide advantages:

  * groups data by a covariate ("patient ID")
  * can be easier to manage (each column one measurement type)



## Converting from wide to long:

```{r}
wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
pivot_longer(wide, cols=c("control", "cond1", "cond2"), 
  names_to="condition", values_to="measurement")
```



## Converting from long to wide

```{r}
long <- read.table(header=TRUE, text='
 subject  sampleID sex condition measurement
       1  ID000001 M   control         7.9
       1  ID000002 M     cond1        12.3
       1  ID000003 M     cond2        10.7
       2  ID000004 F   control         6.3
       2  ID000005 F     cond1        10.6
       2  ID000006 F     cond2        11.1
       3  ID000007 F   control         9.5
       3  ID000008 F     cond1        13.1
       3  ID000009 F     cond2        13.8
')
```

## Converting from long to wide

```{r}
## not what we wanted!!! Why?
pivot_wider(long, names_from="condition", values_from="measurement")

## Instead: 
pivot_wider(long, id_cols="subject", names_from="condition", values_from="measurement")
```

## Exercise 3.3

Convert the following files to long format:

 * `iris.tsv`
 * `labresults_wide.csv`
 * `cars.xlsx` (tricky!)

Clean up and convert to long format (what seems to be the problem? How do
we deal with that?):

 * `mtcars_wide.csv`

